Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 6: refine.

%There were a number of papers categorized under the Communities banner and so
%two presenters each summarized half of the submissions in this area.
Across all talks and papers submitted, authors implicitly and
explicitly recognized the concept of ``communities'' as a driving and
unifying force within software projects.  Despite that, the actual
nature of the communities, the incentive structures and bind them
together, the infrastructure they utilize for communication and even
the types of individuals that make up those communities were
different.  Some of these communities were primarily composed of
members of industry, some were funded and driven by individuals
focused on developing software as opposed to utilizing it, and others
were primarily composed of scientist practitioners, and were true
communities of practice.

@@

Drawing on experiences from high-energy physics, Vay et
al.~\cite{Vay_WSSSPE} proposed developing teams of technical
specialists able to overcome a lack of coordination between projects.
Maximizing scientific output requires maximizing the usability of the
scientific codes while minimizing the cost of developing and
supporting those codes.  This included targeting different
architectures for their software to be deployed, as well as
coordination between technically-focused individuals and usage of a
common scripting language between projects.  Instead of fragmenting
the development of simulation codes across institutions, the paper
suggests that a cohesive strategy reducing duplication and increasing
coordination will broadly increase the efficiency across institutions.
The approach proposed is of de-fragmenting the existing ecosystem in a
non-disruptive way.

@@

Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci
project in developing collaborations supporting tool development for
Open Science.  This software collective, organized around the
statistical programming environment R, develops access mechanisms for
data repositories and attempts to reduce the barrier to entry for
individuals wanting to access data repositories and study the data
contained therein.  The collective fosters direct collaboration
between individuals and data providers, designed to ``train academics
in reproducible science workflows focused around R.''  Two central
challenges are engagement of existing users within ecology and
evolutionary biology (EEB), and how the community can make inroads and
traction in other disciplines. Currently, the collective is exploring
addressing these challenges through use of social media, holding
workshops, and hackathons.  This helps to both raise the profile of
the collective within EEB and in other domains.  However, the
overarching challenge identified in the paper was that of
incentivizing maintenance of software, which is difficult in academia.

@@

Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree
to which research relies on high quality software. There are often
barriers and a lack of suitable incentives for researchers to embrace
software engineering principles. The Water Science Software Institute
is working to lower some of these barriers through an Open Community
Engagement Process. This is a 4-step iterative development process
that incorporates Agile development principles.

@@

Christopherson reports on the application of Steps 1-3 to a
computational modeling framework developed in the 1990s. Step 1 was a
2-day, in-person specifications meeting and code walk-through. Step 2
was a 5-day hackathon to develop working code, and Step 3 was a 3-day
hackathon to refine the code based on new requirements. The team
worked on small, low-risk units of code. It was challenging, revealed
unanticipated obstacles, programmers had to work together, and
experimentation was encouraged.
%
The paper recommended: start small and gradually building toward more
complex objectives, consistent with Agile development; develop
consensus before coding, by repeating step 1 before all hackathons;
ensure newcomers receive orientation prior to the hackathon, such as a
code walk-through or system documentation; and co-locate collaborators
whenever feasible.

@@

Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can
provide a user-friendly entry to complex cyberinfrastructure. For
example, more than 7,000 biologists have run phylogenetic codes on
supercomputers using the \toolname{CIPRES} Science Gateway in 3 1/2
years. Over 120 scientists from 50 institutions used the
\toolname{UltraScan Science Gateway} in one year, increasing the
sophistication of analytical ultracentrifugation experiments
worldwide. The new Neuroscience Gateway (\toolname{NSG}) registered
100+ users who used 250,000 CPU hours in only a few months.

@@

Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and
challenges with the large \toolname{nanoHUB.org} community, where
community is defined as a ``body of persons of common and especially
professional interests scattered through a larger society.'' Support
is challenging because of the diversity of viewpoints and needs. The
group constantly examines its policies to determine whether they are
indirectly alienating part of the community or encouraging particular
types of use.
%
\toolname{nanoHUB}'s 10-year history with over 260,000 users annually
provides a lot of data to analyze. 4000 resources contributed by 1000
authors. \toolname{nanoHUB} serves both the research and education
community and the contribution model allows researchers to get their
codes out into the community and in use in education very rapidly. The
primary software challenges are twofold --- support for the
\toolname{HUBzero} framework and challenges related to the software
contributed by the community.

@@

L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
which was started in 1996 by participants in USA Binary Black Hole
Alliance Challenge. Cactus has a flesh (core) and thorns (modules)
model, a community-oriented framework that allows researchers to
easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain
dormant, becoming active only when parameters and simulation data
dictate. Users can use modules written by others or can write their
own modules without changing other code. The community has grown and
diversified beyond the original science areas.

The paper points out 4 keys to sustaining the community: modular
design, growing a collaborative community, career paths and credit. In
modular design, the Cactus project went far beyond standard practices
of APIs. Domain specific languages (DSLs) allow decoupling of
components --- for example I/O, mesh refinement, PAPI counters, and
boundary conditions abstracted from science code. In academia,
publications are the main currency of credit. Because the project
connects code developments to science, the work is publishable and
modules are citable. Because of the open source, modular approach,
programmers can see the impact of their contributions and often
continue work after graduation. Career paths remain a challenge,
however. Tasks that are essential from a software engineering
perspective are often not rewarded in academia. The best programmers
in a science environment often have multidisciplinary expertise. This
also is not rewarded in academia.

Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF
software institute effort to build a community of those creating
science gateways for science. These gateways, as described in some of
the other papers in this section, can be quite capable and can have
strong scientific impact.  Challenges are similar to those highlighted
by other papers in this section: the conflict between funding for
research vs infrastructure and the challenges around getting academic
credit for infrastructure.
%
The authors observe that development is often done in an isolated
hobbyist environment. Developers are unable to take advantage of
similar work done by others, isolation even when projects have common
needs. But often projects struggle for sustainable funding because
they provide infrastructure to conduct research and many times only
the research is funded. Gateways also may start as a small group
research project, taking off in popularity once they go live, without
any long term plans for sustainability.  Subsequent disruptions in
service can limit effectiveness and test the limits of the research
community's trust. %The impact of gateways can be increased
substantially if we understand
%what makes them successful.

Recommendations from an early study of successful gateways include: 1)
Leadership and management teams should design governance to represent
multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the
technical and domain-related issues, consider sustainability and
measure success early and often. 2) Projects should hire a team of
professionals, demonstrate credibility through stability and clarity
of purpose, leverage the work of others, and plan for flexibility. 3)
Projects should identify an existing community and understand the
communities' needs before beginning, then adapt as the communities'
needs evolve. 4) Funders, should consider the technology project
lifecycle, and design solicitations to reward effective planning,
recognize the benefits and limitations of both technology innovation
and reuse, expect adjustments during the production process, copy
effective models from other industries and sectors, and encourage
partnerships that support gateway sustainability.

@@

\subsubsection{Measuring community}
How might a project know when it has built a sustainable community?
How might an outsider be able to assess the activity and
sustainability of that community?  These questions have been partially
addressed in the literature.  For example, a key metric in online
communities in general is the cross-over point where there are more
external contributors than internal ones. Richard Millington's book
``Buzzing Communities'' does an excellent job of outlining these
measures, drawing on online communities research in an accessible
manner~\cite{millington_buzzing_2012}. Some participants in the
workshop have since prepared online materials outlining current and
future practices in measurement of scientific software and its
ecosystem~\cite{metrics-web}.
%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please
%change to cite}

@@

From more traditional open source development, resources were shared
that developed communities explicitly, such
as~\cite{citeulike:7888211} and~\cite{Trapani:2011}, focusing on
large-scale projects such as the Ubuntu Linux distribution and
smaller-scale volunteer-developed projects like such as ThinkUp,
respectively. The process of open source development, while less
explicitly focused on community building, outlined
in~\cite{citeulike:478633} was seen as a valuable resource,
particularly when combined with the management and personal
interaction techniques outlined in~\cite{opac-b1134063}.  Growing
diversity in communities was directly addressed
in~\cite{Allsopp:2012}, where experiences growing the diversity of
technical conferences in open source were described.

>>>>

%There were a number of papers categorized under the Communities banner and so
%two presenters each summarized half of the submissions in this area.
Across all talks and papers submitted, authors implicitly and
explicitly recognized the concept of ``communities'' as a driving and
unifying force within software projects.  Despite that, the actual
nature of the communities, the incentive structures that bind them
together, the infrastructure they utilize for communication and even
the types of individuals that make up those communities were
different.  Some of these communities were primarily composed of
members of industry, some were funded and driven by individuals
focused on developing software as opposed to utilizing it, and others
were primarily composed of scientist practitioners, and were true
communities of practice.

@@

Drawing on experiences from high-energy physics, Vay et
al.~\cite{Vay_WSSSPE} proposed developing teams of technical
specialists   to overcome a lack of coordination between projects.
Maximizing scientific output requires maximizing the usability of the
scientific code while minimizing the cost of developing and
supporting those code.  This included targeting different
architectures for their software to be deployed, as well as
coordination between technically-focused individuals and usage of a
common scripting language between projects.  Instead of fragmenting
the development of simulation codes across institutions, the paper
suggests that a cohesive strategy reducing duplication and increasing
coordination will broadly increase the efficiency across institutions.
The approach proposed is of de-fragmenting the existing ecosystem in a
non-disruptive way.

@@

Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci
project in developing collaboration supporting tools for
Open Science.  This software collective, organized around the
statistical programming environment R, develops access mechanisms for
data repositories and attempts to reduce the barrier to entry for
individuals wanting to access data repositories and study the data
contained therein.  The collective fosters direct collaboration
between individuals and data providers, designed to ``train academics
in reproducible science workflows focused around R.''  Two central
challenges are engagement of existing users within ecology and
evolutionary biology (EEB), and how the community can make inroads and
traction in other disciplines. Currently, the collective is exploring
addressing these challenges through the use of social media, holding
workshops and hackathons.  This helps to both raise the profile of
the collective within EEB and in other domains.  However, the
overarching challenge identified in the paper was that of
incentivizing maintenance of software, which is difficult in academia.

@@

Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree
to which research relies on high quality software. There are often
barriers and a lack of suitable incentives for researchers to embrace
software engineering principles. The Water Science Software Institute
is working to lower some of these barriers through an Open Community
Engagement Process. This is a four-step iterative development process
that incorporates Agile development principles.

@@

Christopherson reports on the application of Steps~1-3 to a
computational modeling framework developed in the 1990s. Step~1 was a
2-day, in-person specifications meeting and code walk-through. Step 2
was a 5-day hackathon to develop working code, and Step 3 was a 3-day
hackathon to refine the code based on new requirements. The team
worked on small, low-risk units of code. It was challenging, revealed
unanticipated obstacles, programmers had to work together, and
experimentation was encouraged.
%
The paper recommended: start small and gradually building toward more
complex objectives, consistent with Agile development; develop
consensus before coding, by repeating step~1 before all hackathons;
ensure newcomers receive orientation prior to the hackathon, such as a
code walk-through or system documentation; and co-locate collaborators
whenever feasible.

@@

Pierce et al.~\cite{Pierce2_WSSSPE} describes how science gateways can
provide a user-friendly entry to complex cyberinfrastructure. For
example, more than 7,000 biologists have run phylogenetic codes on
supercomputers using the CIPRES Science Gateway in $3 \tfrac{1}{2}$
years. Over 120 scientists from 50 institutions used the
\toolname{UltraScan Science Gateway} in one year, increasing the
sophistication of analytical ultracentrifugation experiments
worldwide. The new Neuroscience Gateway (\toolname{NSG}) registered
more than 100 users who used 250,000 CPU hours in only a few months.

@@

Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and
challenges with the large \toolname{nanoHUB.org} community, where
community is defined as a ``body of persons of common and especially
professional interests scattered through a larger society.'' Support
is challenging because of the diversity of viewpoints and needs. The
group constantly examines its policies to determine whether they are
indirectly alienating part of the community or encouraging particular
types of use.
%
\toolname{nanoHUB}'s 10-year history with over 260,000 users annually
provides a lot of data to analyze: 4000 resources contributed by 1000
authors. \toolname{nanoHUB} serves both the research and education
community and the contribution model allows researchers to get their
codes out into the community and in use in education very rapidly. The
primary software challenges are twofold --- support for the
\toolname{HUBzero} framework and challenges related to the software
contributed by the community.

@@

L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
which was started in 1996 by participants in the USA Binary Black Hole
Alliance Challenge. Cactus has a flesh (core) and thorns (modules)
model, a community-oriented framework that allows researchers to
easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain
dormant, becoming active only when parameters and simulation data
dictate. Users can use modules written by others or can write their
own modules without changing other code. The community has grown and
diversified beyond the original science areas.

The paper points out four keys to sustaining the community: modular
design, growing a collaborative community, career paths, and credit. In
modular design, the Cactus project went far beyond standard practices
of APIs. Domain specific languages (DSLs) allow decoupling of
components --- for example I/O, mesh refinement, PAPI counters, and
boundary conditions abstracted from science code. In academia,
publications are the main currency of credit. Because the project
connects code developments to science, the work is publishable and
modules are citable. Because of the open source, modular approach,
programmers can see the impact of their contributions and often
continue work after graduation. Career paths remain a challenge,
however. Tasks that are essential from a software engineering
perspective are often not rewarded in academia. The best programmers
in a science environment often have multidisciplinary expertise. This
also is not rewarded in academia.

Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF
software institute effort to build a community of those creating
science gateways for science. These gateways, as described in some of
the other papers in this section, can be quite capable and can have
strong scientific impact.  Challenges are similar to those highlighted
by other papers in this section: the conflict between funding for
research vs infrastructure and the challenges around getting academic
credit for infrastructure.
%
The authors observe that development is often done in an isolated
hobbyist environment. Developers are unable to take advantage of
similar work done by others, finding themselves in isolation even when their projects have common
goals. But often projects struggle for sustainable funding because
they provide infrastructure to conduct research and many times only
the research is funded. Gateways also may start as a small group
research project, taking off in popularity once they go live, without
any long term plans for sustainability.  Subsequent disruptions in
service can limit effectiveness and test the limits of the research
community's trust. %The impact of gateways can be increased
%substantially if we understand
%what makes them successful.

Recommendations from an early study of successful gateways include: 1)
Leadership and management teams should design governance to represent
multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the
technical and domain-related issues, consider sustainability and
measure success early and often. 2) Projects should hire a team of
professionals, demonstrate credibility through stability and clarity
of purpose, leverage the work of others, and plan for flexibility. 3)
Projects should identify one ore more existing communities and understand the
communities' needs before beginning, then adapt as the communities'
needs evolve. 4) Funders, should consider the technology project
lifecycle, and design solicitations to reward effective planning,
recognize the benefits and limitations of both technology innovation
and reuse, expect adjustments during the production process, copy
effective models from other industries and sectors, and encourage
partnerships that support gateway sustainability.

@@

\subsubsection{Measuring community}
How might a project know when it has built a sustainable community?
How might an outsider be able to assess the activity and
sustainability of that community?  These questions have been partially
addressed in the literature.  For example, a key metric in online
communities in general is the cross-over point where there are more
external contributors than internal ones. Richard Millington's book
``Buzzing Communities'' does an excellent job of outlining these
measures, drawing on communities research in an accessible
manner~\cite{millington_buzzing_2012}. Some participants in the
workshop have since prepared materials outlining current and
future practices in measurement of scientific software and its
ecosystem~\cite{metrics-web}.
%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please
%change to cite}

@@

From more traditional open source development, resources were shared
that developed communities explicitly, such
as~\cite{citeulike:7888211} and~\cite{Trapani:2011}, focusing on
large-scale projects such as the Ubuntu Linux distribution and
smaller-scale volunteer-developed projects like such as ThinkUp,
respectively. The process of open source development, while less
explicitly focused on community building, sketched
in~\cite{citeulike:478633} was seen as a valuable resource,
particularly when combined with the management and personal
interaction techniques outlined in~\cite{opac-b1134063}.  Growing
diversity in communities was directly addressed
in~\cite{Allsopp:2012}, where experiences growing the diversity of
technical conferences in open source were described.
